### 2020-12-01 ch02_scientific_background/scientific_background.tex
Added first part of carbon cycle section

Due to the complex nature of the Earth system itself, numerical models of it
consist of hundreds of thousands of lines of computer code. Thus, a
standardization to a certain degree is crucial for the various research groups
developing \acp{ESM} all around the world in order to obtain comparable output
and to facilitate analysis. For this reason, the \ac{WGCM} of the \ac{WCRP}
initiated the \ac{CMIP} in 1995, with the objective to \enquote{better
understand past, present and future climate changes arising from natural,
unforced variability or in response to changes in radiative forcing in a
multi-model context} \autocite{WCRP2020}. One major element of \ac{CMIP} is to
establish common standards, coordination, infrastructure, and documentation in
order to facilitate the distribution of climate model output
\autocite{Eyring2016}.

@@

To increase diversity and answer more scientific questions, \acs{CMIP}6 models
can participate in the so-called \acs{CMIP}6-Endorsed \acp{MIP}, of which
\acs{CMIP}6 offers 21 (see circular sectors in \cref{fig:02:cmip6}). Some
\acp{MIP} offer additional experiments to explore specific aspects of the Earth
system, like the \ac{C4MIP} which focuses on the carbon cycle
\autocite{Jones2016} or the \ac{CFMIP} which focuses on the evaluation of cloud
feedbacks \autocite{Webb2017}. Other \acp{MIP} allow the assessment of future
climate change. An example is the \ac{ScenarioMIP}, which provides common
experiments that simulate different possible futures \autocite{ONeill2016}.
These experiments are based on the so-called \acp{SSP}, a set of alternative
pathways of future societal development \autocite{ONeill2013, ONeill2017}. For
each experiment, a set of emissions and land use changes is calculated from the
\acp{SSP} \autocite{Riahi2017} which are then used to force the global climate
models. For \ac{ScenarioMIP}, five different \acp{SSP} are considered, ranging
from \ac{SSP}1 (sustainability) to \ac{SSP}5 (fossil-fuel development). Each
\ac{SSP} is combined with a climate outcome (measured as radiative forcing in
the year 2100) based on a particular forcing pathway that \acp{IAM} have shown
to be feasible. For example, \ac{SSP}5-8.5 represents a scenario based on a
fossil-fuel development with a radiative forcing of $8.5 \unit{W m^{-2}}$ in
2100 while \ac{SSP}1-2.6 represents a sustainable future with a radiative
forcing of $2.6 \unit{W m^{-2}}$ in the year 2100. The two other main scenarios
(called \enquote{Tier 1} experiments in \ac{ScenarioMIP}) are the \ac{SSP}2-4.5
and \ac{SSP}3-7.0 scenarios. In contrast to the \ac{ScenarioMIP} experiments,
the corresponding \acs{CMIP}5 counterparts \autocite{Taylor2012}, the so-called
\ac{RCP}, only used the radiative forcing in 2100 as only dimension to describe
the possible futures (\eg{} \ac{RCP}8.5, \ac{RCP}4.5, \ac{RCP}2.6, \etc{}).

@@

@@

I like \ac{GPP}!

@@

>>>>

Due to the complex nature of the Earth system itself, numerical models of it
consist of hundreds of thousands of lines of computer code. Thus, a
standardization to a certain degree is crucial for the various research groups
developing \acp{ESM} all around the world in order to obtain comparable output
and to facilitate model analysis. For this reason, the \ac{WGCM} of the
\ac{WCRP} initiated the \ac{CMIP} in 1995, with the objective to
\enquote{better understand past, present and future climate changes arising
  from natural, unforced variability or in response to changes in radiative
  forcing in a multi-model context} \autocite{WCRP2020}. One major element of
\ac{CMIP} is to establish common standards, coordination, infrastructure, and
documentation in order to facilitate the distribution of climate model output
\autocite{Eyring2016}.

@@

To increase diversity and answer more scientific questions, \acs{CMIP}6 models
can participate in the so-called \acs{CMIP}6-Endorsed \acp{MIP}, of which
\acs{CMIP}6 offers 21 (see circular sectors in \cref{fig:02:cmip6}). Some
\acp{MIP} offer additional experiments to explore specific aspects of the Earth
system, like the \ac{C4MIP} which focuses on the carbon cycle
\autocite{Jones2016} or the \ac{CFMIP} which focuses on the evaluation of cloud
feedbacks \autocite{Webb2017}. Other \acp{MIP} allow the assessment of future
climate change. An example is the \ac{ScenarioMIP}, which provides common
experiments that simulate different possible futures \autocite{ONeill2016}.
These experiments are based on the so-called \acp{SSP}, a set of alternative
pathways of future societal development \autocite{ONeill2013, ONeill2017}. For
each experiment, a set of emissions and land use changes is calculated from the
\acp{SSP} \autocite{Riahi2017} which are then used to force the global climate
models. For \ac{ScenarioMIP}, five different \acp{SSP} are considered, ranging
from \ac{SSP}1 (sustainability) to \ac{SSP}5 (fossil-fuel development). Each
\ac{SSP} is combined with a climate outcome (measured as radiative forcing in
the year 2100) based on a particular forcing pathway that \acp{IAM} have shown
to be feasible. For example, \ac{SSP}5-8.5 represents a scenario based on a
fossil-fuel development with a radiative forcing of $8.5 \unit{W m^{-2}}$ in
2100 while \ac{SSP}1-2.6 represents a sustainable future with a radiative
forcing of $2.6 \unit{W m^{-2}}$ in the year 2100. The two other main scenarios
(called \enquote{Tier 1} experiments in \ac{ScenarioMIP}) are the \ac{SSP}2-4.5
and \ac{SSP}3-7.0 scenarios. In contrast to the \ac{ScenarioMIP} experiments,
the corresponding \acs{CMIP}5 counterparts \autocite{Taylor2012}, the so-called
\acp{RCP}, only used the radiative forcing in 2100 as only dimension to
describe the possible futures (\eg{} \ac{RCP}8.5, \ac{RCP}4.5, \ac{RCP}2.6,
\etc{}).

@@

- anthropogenic forcing vs. natural forcing (volcanoes, solar cycle, etc.)

- basic physics of greenhouse gases (vibration modes, etc.)

- historical vs. future: in historical (future), uncertainty in forcing low
(high)

- definition of feedbacks, examples

\section{Climate Sensitivity}
\label{sec:02:climate_sensitivity}

- Physical Feedbacks (ECS)

- Equations for feedbacks + decomposition

- ocean heat uptake (TCR)

- different between effective climate sensitivity and equilibrium version

- equations for ECS and TCR

@@

Since one study presented in this thesis \autocite[see
\cref{ch:06:paper_gpp}]{Schlund2020} aims to reduce uncertainties in carbon
cycle-related processes, this chapter introduces the scientific background of
the global carbon cycle.

\subsection{Overview}
\label{subsec:02:carbon_cycle_overview}

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidth{}]{
    ch02_scientific_background/figs/carbon_cycle_schematic.png}
  \caption{Simplified schematic of the global carbon cycle including the
    typical turnover time scales for carbon transfers through the major
    reservoirs (atmosphere, land surface and ocean). Taken from
    \textcite{Ciais2013}.}
  \label{fig:02:carbon_cycle_schematic}
\end{figure}

A schematic overview of the global carbon cycle is shown in
\cref{fig:02:carbon_cycle_schematic}. To quantify the carbon cycle, common
units are \ac{ppm} for the atmospheric trace gas concentrations (dry-air mole
fraction) and \ac{GtC} or $\si{\GtCyr{}}$ for the reservoirs masses or exchange
fluxes, respectively. The carbon exchange processes between the different
carbon reservoirs run on a wide range of time scales. Conceptually, one can
distinguish between two domains of the global carbon cycle: a slow and a fast
domain. The slow domain with turnover times (reservoir mass of carbon divided
by exchange flux) of more than 10000 years consists of the large carbon stores
in rocks and sediments which are connected to the rapid domain of the carbon
cycle through volcanic emissions of \ac{CO2}, chemical weathering, erosion and
sediment formation on the sea floor. These natural exchange fluxes between the
slow and the fast domain are comparatively small ($< 0.3 \unit{\GtCyr{}}$) and
can be assumed as approximately constant in time over the last few centuries
\autocite{Ciais2013}.

The fast domain of the global carbon cycle consists of three main carbon
reservoirs: the atmosphere, the terrestrial biosphere and the ocean. In the
atmosphere, carbon is mainly stored in trace gases, with \ac{CO2} as the major
component with a current (2019) concentration of about $410 \unit{ppm}$
\autocite{Friedlingstein2019}. Additional contributors to the atmospheric
carbon content are the trace gas \ac{CH4}, the trace gas \ac{CO}, hydrocarbons,
black carbon aerosols and organic compounds \autocite{Ciais2013}. Carbon in the
terrestrial biosphere is mainly stored as organic compounds, with about
$\range{450}{650} \unit{GtC}$ in the living vegetation biomass,
$\range{1500}{2400} \unit{GtC}$ in dead organic matter in litter and soils and
about $1700 \unit{GtC}$ in permafrost soils \autocite{Ciais2013}. The main
component of the oceanic carbon reservoir is dissolved inorganic carbon
(carbonic acid, bicarbonate ions and carbonate ions) with about $38000
\unit{GtC}$. Further carbon is stored as dissolved organic carbon (about $700
\unit{GtC}$), in surface sediments (about $1750 \unit{GtC}$) and in marine
biota (about $3 \unit{GtC}$, predominantly phytoplankton and other
microorganisms) \autocite{Ciais2013, Friedlingstein2019}.

In the fast domain of the global carbon cycle, reservoir turnover times range
from seconds to millennia. In contrast to the slow domain, the carbon exchange
fluxes within the fast domain of the carbon cycle are much higher. One major
group of exchange processes in the fast domain connects the atmosphere and the
terrestrial biosphere. \ac{CO2} is removed from the atmosphere by plant
photosynthesis with about $120 \unit{\GtCyr{}}$ \autocite{Ciais2013}. This
process is also known as \ac{GPP}. The carbon fixed into plants can be released
back into the atmosphere by autotrophic (plant) and heterotrophic (soil
microbial and animal) respiration and additional disturbance processes like
fires \autocite{Ciais2013}. Since the land \ac{CO2} uptake by photosynthesis
occurs only during the growing season, whereas respiration occurs nearly all
year, the larger amount of vegetation in the Northern hemisphere (due to the
larger land mass) gives rise to a seasonal cycle of the atmospheric \ac{CO2}
concentration \autocite{Keeling1995}. This seasonal cycle reflects the phase of
the global carbon cycle and shows a maximum of the atmospheric \ac{CO2}
concentration in the Northern hemisphere winter (net \ac{CO2} flux into
atmosphere due to respiration) and a minimum during the Northern hemisphere
summer (net \ac{CO2} flux into the land due to photosynthesis). Another major
carbon exchange process connects the atmosphere and the ocean. Atmospheric
\ac{CO2} is exchanged with the surface ocean through gas exchange, which is
driven by the partial \ac{CO2} pressure difference between the air and the sea
\autocite{Ciais2013}.

\begin{figure}[t]
  \centering
  \includegraphics[width=\largeFigureWidth{}]{
    ch02_scientific_background/figs/carbon_cycle_perturbation.png}
  \caption{Schematic representation of the overall perturbation of the global
    carbon cycle caused by anthropogenic activities, averaged globally for the
    decade \range{2009}{2018}. Arrows represent carbon exchange fluxes; circles
    carbon reservoirs. More details are given in the legend of this figure.
    Taken from \textcite{Friedlingstein2019}.}
  \label{fig:02:carbon_cycle_perturbation}
\end{figure}

@@

- Process-based weighting (Knutti + MDER)

- emergent constraints

- Wenzel et al., Nature (2016) as example (refer to carbon cycle chapter)

- Review of ECS emergent constraints is given in chapter on Schlund et al., ESD
(2020)
